hormones

### 2019-11-11 analysis/07_hormones.Rmd
removal and timeing in two differnt plots

source("../R/functions.R")  # load custom functions 
source("../R/themes.R")  # load custom themes and color palletes

@@

```{r clean}
charlevels <- c("control", "bldg", "lay", "inc_d3", "inc_d9", "inc_d17", "hatch", "n5", "n9" )
removelevels <- c("M_inc3",  "M_inc9", "M_inc17", "M_n2")
maniplevels <- c("early", "prolong", "extend")
combolevels <- c("control", "bldg", "lay", "inc_d3", "M_inc3",
                  "inc_d9",  "M_inc9", "early",  "inc_d17","M_inc17",  "hatch",
                 "prolong", "extend", "M_n2",
                  "n5", "n9" )

manipchar <- c("early",  "inc_d17", "hatch", "prolong", "extend", "n5")
removechar <- c("inc_d3", "M_inc3",  "inc_d9", "M_inc9","inc_d17", "M_inc17","hatch", "M_n2")

sexcolors <- c("female" = "#F8766D", "male" = "#00BFC4")
charcolors <- c("control" = "#F8766D", "bldg" = "#D39200",
                "lay" =  "#93AA00", "inc_d3" = "#00BA38" ,
                "inc_d9" = "#00C19F" , "inc_d17" =  "#00B9E3", 
                "hatch" = "#619Cff" , "n5" =  "#DB72Fb", "n9" =  "#FF61C3")

charsexlevels <- c("control female", "control male", "bldg female", "bldg male",  "lay female", "lay male",
                "inc_d3 female", "inc_d3 male", "inc_d9 female", "inc_d9 male",
                "inc_d17 female", "inc_d17 male", "hatch female", "hatch male", 
                "n5 female", "n5 male", "n9 female" , "n9 male")

@@

# rename some of the timepoints levels, store in new columns
prolactin <- prolactin %>%
    mutate(sex = fct_recode(Sex,
                            "female" = "f",
                            "male" = "m"),
           timepoint = fct_recode(Treatment,
                            "hatch" = "Hatch",
                            "inc_d17" = "Inc_d17",
                            "inc_d3" = "Inc_d3",
                            "inc_d9" = "Inc_d9",
                            "M_inc9" = "M_Inc9",
                            "M_inc3" = "M_Inc3",
                            "early" = "M_Inc8",
                            "early" = "M_inc8",
                            "M_inc17" = "M_Inc17",
                            "M_n2" = "M_hatch",
                            "control" = "baseline",
                            "n5" = "N5", 
                            "n9" = "N9"),
           study = fct_collapse(timepoint,
                                 characterization = charlevels,
                                 manipulation = maniplevels,
                                removal = removelevels))

colnames(prolactin)[colnames(prolactin)=="Prolactin ng/mL"] <- "plasma_conc"
prolactin$hormone <- "prolactin"
prolactin <- prolactin %>% select(study, timepoint, sex, hormone, plasma_conc)   %>% drop_na()

@@

PETC <- PETC %>% select(stage, sex, hormone, plasma_conc)  %>%
                filter(stage %in% combolevels)  %>%
                droplevels()  
PETC$stage <- factor(PETC$stage)
levels(PETC$stage)

PETC <- PETC %>%
    mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
           study = fct_collapse(stage,
                                characterization = charlevels,
                                manipulation = maniplevels)) %>% 
  drop_na()
colnames(PETC)[colnames(PETC)=="stage"] <- "timepoint"
PETC <- PETC %>% select(study, timepoint, sex, hormone, plasma_conc)
summary(PETC)
```

```{r combine-clean-split}
hormones <- rbind(prolactin, PETC)
hormones$timepoint <- factor(hormones$timepoint, levels = combolevels)

hormones <- hormones %>%  mutate(timepoint.sex = paste(timepoint, sex, sep = " ")) 
hormones$timepoint.sex <- factor(hormones$timepoint.sex, levels = charsexlevels)

hormones$okay <- ifelse(hormones$hormone == "cort" & hormones$plasma_conc > 30, "bad",
                    ifelse(hormones$hormone == "progesterone" & hormones$plasma_conc > 5, "bad", 
                           ifelse(hormones$hormone == "prolactin" & hormones$plasma_conc > 150, "bad", 
                        ifelse(hormones$hormone == "testosterone" & hormones$sex == "female", "bad",
                               ifelse(hormones$hormone == "estradiol" & hormones$sex == "male", "bad", "okay")))))
hormones <- hormones %>% filter(okay == "okay") %>% droplevels()
summary(hormones)

prl.char <- hormones %>% filter(hormone == "prolactin", timepoint %in% charlevels)   %>%  droplevels()
test.char <- hormones %>% filter(hormone == "testosterone", timepoint %in% charlevels)   %>%  droplevels()
est.char <- hormones %>% filter(hormone == "estradiol", timepoint %in% charlevels)   %>%  droplevels()
prog.char <- hormones %>% filter(hormone == "progesterone", timepoint %in% charlevels)   %>%  droplevels()
cort.char <- hormones %>% filter(hormone == "cort", timepoint %in% charlevels)   %>%  droplevels()

@@

  mycolors <- charcolors
  sexcolors <- c("female" = "#969696", "male" = "#525252")

  hormones %>% 
    filter(study == "characterization",
           hormone %in% c(myhormone))  %>% 
    droplevels() %>% 
  ggplot(aes(x = as.numeric(timepoint), y = plasma_conc)) +
        geom_smooth(aes(colour = sex)) +
    geom_boxplot(aes(outlier.colour = timepoint, fill = timepoint, alpha = sex)) +
    theme_B3() +
    theme(axis.text.x =  element_blank(),
          legend.position = "none") +
    scale_fill_manual(values = mycolors) +
    scale_color_manual(values = sexcolors) +
    labs(y = myylab, x = NULL) +
    guides(fill = guide_legend(order=1),
         color = guide_legend(order=2)) +
    scale_alpha_manual(values = c(0.75,1)) +
    scale_color_manual(values = c("female" = "#969696", "male" = "#525252")) 
  }

@@

```{r PRLonly, fig.width = 2.25, fig.height = 1}
hormonecharplot("prolactin", "PRL (ng/mL)")
```

```{r sexsteroids, fig.width=2.25, fig.height= 3}
b <- hormonecharplot("cort", "CORT (ng/mL)")
c <- hormonecharplot("progesterone", "PROG (ng/mL)") 
d1 <- hormonecharplot("estradiol", "E (ng/mL)")
d2 <- hormonecharplot("testosterone", "T (ng/mL)")
d <- plot_grid(d1,d2, nrow = 1)
bcd <- plot_grid(b ,c, d, ncol = 1)
bcd
```

```{r aov-char}
aovSexTretment <- function(mydata, whichormone){
  aov2 <- aov(data = mydata, plasma_conc ~ timepoint + sex)
  print(whichormone)
  print(summary(aov2))
  #print(TukeyHSD(aov2, which = "timepoint"))
}

aovTretment  <- function(mydata, whichormone){
  aov1 <- aov(data = mydata, plasma_conc ~ timepoint )
  print(whichormone)
  print(summary(aov1))
  #print(TukeyHSD(aov1, which = "timepoint"))
}

aovSexTretment(prl.char, "PRL")
aovSexTretment(cort.char, "CORT")
aovSexTretment(prog.char, "PROG")

aovTretment(est.char, "E")
aovTretment(test.char, "T")
```

```{r manipulation}
hormonemanipPRL <- function(myhormone, myylab){

  mycolors <- sexcolors

  hormones %>% 
    filter(timepoint %in% manipchar,
           hormone %in% c(myhormone))  %>% 
  ggplot(aes(x = timepoint, y = plasma_conc, fill = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = mycolors) +
    labs(y = myylab, x = NULL) +
    facet_wrap(~sex) +
    annotate("rect", xmin = 0.5, xmax = 1.5, ymin = 0, ymax = 125, alpha = .2) +
    annotate("rect", xmin = 3.5, xmax = 5.5, ymin = 0, ymax = 125, alpha = .2) 
}

hormonemanipPRL("prolactin", "PRL (ng/mL)")

@@

  mycolors <- sexcolors

  hormones %>% 
    filter(timepoint %in% manipchar,
           hormone %in% c(myhormone))  %>% 
  ggplot(aes(x = timepoint, y = plasma_conc, fill = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = mycolors) +
    labs(y = myylab, x = NULL) +
    facet_wrap(~sex) +
    annotate("rect", xmin = 2.5, xmax = 4.5, ymin = 0, ymax = myymax, alpha = .2)
}

@@

plot_grid(d,a,c,b, rel_widths = c(0.65,0.35))
```
```{r removal}
hormoneremoval <- function(myhormone, myylab){

  mycolors <- c("female" = "#F8766D", "male" = "#00BFC4")

  hormones %>% 
    filter(timepoint %in% removechar,
           hormone %in% c(myhormone))  %>% 
  ggplot(aes(x = timepoint, y = plasma_conc, fill = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = mycolors) +
    labs(y = myylab, x = NULL) +
    facet_wrap(~sex) +
    annotate("rect", xmin = 1.5, xmax = 2.5, ymin = 0, ymax = 125, alpha = .2) +
    annotate("rect", xmin = 3.5, xmax = 4.5, ymin = 0, ymax = 125, alpha = .2) +
    annotate("rect", xmin = 5.5, xmax = 6.5, ymin = 0, ymax = 125, alpha = .2) + 
    annotate("rect", xmin = 7.5, xmax = 8.5, ymin = 0, ymax = 125, alpha = .2) 
}

hormoneremoval("prolactin", "PRL (ng/mL)")
```

>>>>

source("../R/themes.R")  # load custom themes and color palletes

@@

@@

# rename some of the treatments levels, store in new columns
prolactin <- prolactin %>%
    mutate(sex = fct_recode(Sex,
                            "female" = "f",
                            "male" = "m"),
           treatment = fct_recode(Treatment,
                            "hatch" = "Hatch",
                            "inc.d17" = "Inc_d17",
                            "inc.d17" = "inc_d17",
                            "inc.d3" = "Inc_d3",
                             "inc.d3" = "inc_d3",
                            "inc.d9" = "Inc_d9",
                             "inc.d9" = "inc_d9",
                            "m.inc.d9" = "M_Inc9",
                            "m.inc.d9" = "M_inc9",
                            "m.inc.d3" = "M_Inc3",
                            "m.inc.d8" = "M_Inc8",
                            "m.inc.d8" = "M_inc8",
                            "m.inc.d17" = "M_Inc17",
                            "m.n2" = "M_hatch",
                            "control" = "baseline",
                            "n5" = "N5", 
                            "n9" = "N9"),
           study = fct_collapse(treatment,
                                 characterization = charlevels,
                                 manipulation = maniplevels1))

colnames(prolactin)[colnames(prolactin)=="Prolactin ng/mL"] <- "plasma_conc"
prolactin$hormone <- "prolactin"
prolactin <- prolactin %>% 
              select(study, treatment, sex, hormone, plasma_conc)  %>% 
              drop_na()

@@

PETC <- PETC %>% select(stage, sex, hormone, plasma_conc)  %>%
                mutate(treatment = fct_recode(stage,
                            "inc.d17" = "inc_d17",
                            "inc.d3" = "inc_d3",
                            "inc.d9" = "inc_d9",
                            "m.inc.d9" = "m_incd9",
                            "m.inc.d3" = "m_incd3",
                            "m.inc.d8" = "m_incd8",
                            "m.inc.d17" = "m_incd17",
                            "m.n2" = "m_hatch",
                            "control" = "stress_hpg")) %>%
                filter(treatment %in% alllevels2) %>%   
                mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
                       study = fct_collapse(treatment,
                                characterization = charlevels,
                                manipulation = maniplevels1)) %>% 
              drop_na() %>%  droplevels()  %>% 
              select(study, treatment, sex, hormone, plasma_conc)
summary(PETC)
```

```{r combine-clean-split}
hormones <- rbind(prolactin, PETC)
hormones$treatment <- factor(hormones$treatment, levels = alllevels)

hormones$okay <- ifelse(hormones$hormone == "cort" & hormones$plasma_conc > 30, "bad",
                    ifelse(hormones$hormone == "progesterone" & hormones$plasma_conc > 5, "bad", 
                           ifelse(hormones$hormone == "prolactin" & hormones$plasma_conc > 150, "bad", 
                        ifelse(hormones$hormone == "testosterone" & hormones$sex == "female", "bad",
                               ifelse(hormones$hormone == "estradiol" & hormones$sex == "male", "bad", "okay")))))
hormones <- hormones %>% filter(okay == "okay") %>% droplevels()
summary(hormones)
```

  hormones %>% 
    filter(study == "characterization",
           hormone %in% c(myhormone))  %>% 
    droplevels() %>% 
  ggplot(aes(x = as.numeric(treatment), y = plasma_conc)) +
        geom_smooth(aes(colour = sex)) +
    geom_boxplot(aes(outlier.colour = treatment, fill = treatment, alpha = sex)) +
    mytheme() +
    theme(axis.text.x =  element_blank(),
          legend.position = "none") +
    scale_fill_manual(values = colorscharmaip) +
    scale_color_manual(values = sexcolors) +
    labs(y = myylab, x = NULL) +
    guides(fill = guide_legend(order=1),
         color = guide_legend(order=2)) +
    scale_alpha_manual(values = c(0.75,1)) +
    scale_color_manual(values = c("female" = "#969696", "male" = "#525252")) 
  }

@@

```{r PRLonly}
hormonecharplot("prolactin", "PRL (ng/mL)")
```

```{r sexsteroids}
b <- hormonecharplot("cort", "CORT (ng/mL)")
c <- hormonecharplot("progesterone", "PROG (ng/mL)") 
d1 <- hormonecharplot("estradiol", "E (ng/mL)")
d2 <- hormonecharplot("testosterone", "T (ng/mL)")
d <- plot_grid(d1,d2, nrow = 1)
bcd <- plot_grid(b ,c, d, ncol = 1)
bcd
```

```{r aov-char}
prl.char <- hormones %>% filter(hormone == "prolactin", treatment %in% charlevels)   %>%  droplevels()
test.char <- hormones %>% filter(hormone == "testosterone", treatment %in% charlevels)   %>%  droplevels()
est.char <- hormones %>% filter(hormone == "estradiol", treatment %in% charlevels)   %>%  droplevels()
prog.char <- hormones %>% filter(hormone == "progesterone", treatment %in% charlevels)   %>%  droplevels()
cort.char <- hormones %>% filter(hormone == "cort", treatment %in% charlevels)   %>%  droplevels()

aovSexTretment <- function(mydata, whichormone){
  aov2 <- aov(data = mydata, plasma_conc ~ treatment * sex)
  print(whichormone)
  print(summary(aov2))
  #print(TukeyHSD(aov2, which = "treatment"))
}

aovTretment  <- function(mydata, whichormone){
  aov1 <- aov(data = mydata, plasma_conc ~ treatment )
  print(whichormone)
  print(summary(aov1))
  #print(TukeyHSD(aov1, which = "treatment"))
}

aovSexTretment(prl.char, "PRL") # yes, sex difference (p = 0.00256), yes treatment effect (p < 2e-16), no interaction
aovSexTretment(cort.char, "CORT") # no sex difference (p = 0.779 ), small treatment effect (p = 0.0238),  no interaction
aovSexTretment(prog.char, "PROG") # no sex difference or treatment effect, signifiant interacion (p = 0.0107)

aovTretment(est.char, "E") # p = 0.101
aovTretment(test.char, "T") # p = 0.609
```

```{r manipulation}
# prolactin, removal only
hormones %>% 
    filter( hormone == c("prolactin"))  %>% 
    filter( !treatment %in% c("m.inc.d8", "prolong", "extend"))  %>% 
  ggplot(aes(x = treatment, y = plasma_conc, fill = treatment,  color = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = colorscharmaip) +
    scale_color_manual(values = sexcolors) +
    labs(y = "PRL (ng/mL)", x = NULL) +
    annotate("rect", xmin = 3.7, xmax = 5.3, ymin = -2, ymax = 0, alpha = 1) +
    annotate("rect", xmin = 5.7, xmax = 7.3, ymin = -2, ymax = 0, alpha = 1) +
    annotate("rect", xmin = 7.7, xmax = 9.3, ymin = -2, ymax = 0, alpha = 1) +
    annotate("rect", xmin = 9.7, xmax = 11.3, ymin = -2, ymax = 0, alpha = 1)

hormones %>% 
    filter( hormone == c("prolactin"))  %>% 
    filter( !treatment %in% c("m.inc.d3", "m.inc.d9", "m.inc.d17","m.n2"))  %>% 
  ggplot(aes(x = treatment, y = plasma_conc, fill = treatment,  color = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = colorscharmaip) +
    scale_color_manual(values = sexcolors) +
    labs(y = "PRL (ng/mL)", x = NULL) +
    annotate("rect", xmin = 4.7, xmax = 6.3, ymin = -2, ymax = 0, alpha = 1, alpha = 1) +
    annotate("rect", xmin = 6.7, xmax = 8.3, ymin = -2, ymax = 0, alpha = 1, alpha = 1) +
    annotate("rect", xmin = 8.7, xmax = 10.3, ymin = -2, ymax = 0, alpha = 1, alpha = 1)

@@

@@

  hormones %>% 
    filter( hormone %in% c(myhormone))  %>% 
  ggplot(aes(x = treatment, y = plasma_conc, fill = treatment, color = sex)) +
    geom_boxplot() + 
    mytheme() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
         # legend.position = "none",
          strip.text = element_blank()) +
    scale_fill_manual(values = colorscharmaip) +
    scale_color_manual(values = sexcolors) +
    labs(y = myylab, x = NULL) 
}

@@

a
b
c
d
```
